### 2015-05-12 background.tex
Worked with Pantazis on the theory part of the paper and the new proposed structure of the implementation section.

\subsection{Reasoning about Thread Interleavings}
\label{bg:happensbefore}

\ADComment{I am unsure as to whether we need this background, but I'm undecided as I'm still getting on top of the rest of the material.}

A fundamental concept in concurrency analysis is Lamport's \emph{happens-before} relation~\cite{lamport1978time}, which denotes a partial-order between all events during the execution of a concurrent program. Happens-before can be defined as follows:

\begin{definition}
\label{definition:datarace}
Let $E_1$ and $E_2$ be two events during a program's concurrent execution. If $E_1$ and $E_2$ are in the same thread and $E_1$ occurs before $E_2$, or if $E_1$ and $E_2$ are in different threads and there is a synchronization mechanism (e.g. locking) that prevents a thread schedule in which $E_2$ occurs before $E_1$, then $E_1 \rightarrow E_2$, where the symbol $\rightarrow$ denotes the happens-before relation.
\end{definition}

Happens-before is transitive, i.e. for any three events $E_1$, $E_2$ and $E_3$, if $E_1 \rightarrow E_2$ and $E_2 \rightarrow E_3$, then $E_1 \rightarrow E_3$. If two events $E_1$ and $E_2$ occur in different threads, and neither $E_1 \rightarrow E_2$, nor $E_2 \rightarrow E_1$ holds, then $E_1$ and $E_2$ happen concurrently, which indicates a data race if the two events access the same shared resource and at least one of the accesses is a modifier.

\begin{figure}[htbp]
\centering
\includegraphics[width=1\linewidth]{img/happens_before.pdf}
\caption{Applying happens-before in two concurrently running threads}
\label{fig:happens_before}
\end{figure}

Figure~\ref{fig:happens_before} shows an example of applying happens-before reasoning in two concurrently running entry points. Each of the three statements $\{A, B, C\}$ in thread 1 is executed before the next one, and thus all of them are ordered by the happens-before relation. The same applies for each of the four statements $\{D, E, F, G\}$ in thread 2. Statement $C$ in thread 1 releases the lock, and subsequently allows $D$ in thread 2 to acquire the same lock. Thus, $C \rightarrow D$. This means that for this specific thread interleaving the set of statements $\{A, B, C\}$ happens before the set of statements $\{D, E, F, G\}$.

The above example also showcases the main limitation of happens-before. If the technique explores \emph{only} this specific thread interleaving, it will miss a potential data race between statements $B$ and $G$, because $B \rightarrow G$ in this specific schedule. Thus, a tool based on happens-before requires a scheduler that produces as many different thread interleavings as possible to increase execution path coverage~\cite{savage1997eraser}. Achieving full coverage with tools based on happens-before is infeasible though, because of the exponential number of possible thread interleavings in realistic concurrent programs. Moreover, attempting to explore a large number of thread schedules quickly leads to scalability issues~\cite{musuvathi2008finding}, a problem known as state space explosion. Thus, previous work based on happens-before typically explore paths under a certain bound, such as the number of allowed context switches~\cite{qadeer2004kiss}.

\subsection{Reasoning about Locksets}
\label{bg:lockset}

An alternative approach to happens-before, known as \emph{lockset analysis}, resolves around reasoning about the locking behavior of concurrent programs. Lockset analysis was proposed in the context of Eraser~\cite{savage1997eraser}, a dynamic data race detector that tracks the set of locks that \emph{consistently} protect a memory location during program execution. If that lockset ever becomes empty, the analysis reports a potential data race on that memory location. \ADComment{This could be better explained, I think.} This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.

@@

In contrast to analyses based on happens-before, lockset analysis has the potential to scale well, because it does not need to reason about all possible thread interleavings to achieve good code coverage. The technique, though, suffers from imprecision (i.e. can report false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, the code coverage in dynamic lockset analyzers, such as Eraser, is limited by the execution paths that are explored under a given scheduler. This is because the tool runs the program in a truly concurrent environment, thus it only gets to see what the thread did on the particular schedule that was followed.

>>>>

\subsection{Lockset analysis}
\label{bg:lockset}

Lockset analysis is a lightweight race detection method proposed in the context of Eraser~\cite{savage1997eraser}, a dynamic data race detector.  The idea is to track the set of locks that are \emph{consistently} used to protect a memory location during program execution. If that lockset ever becomes empty, the analysis reports a potential data race on that memory location. This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.

@@

In contrast to more sophisticated race analyses that encode a \emph{happens-before} relation between threads~\cite{TODO} (e.g.\ using vector clocks), lockset analysis is easy to implement, lightweight, and thus has the potential to scale well.  The technique, though, suffers from imprecision (i.e.\ can report false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, the code coverage in dynamic lockset analyzers, such as Eraser, is limited by the execution paths that are explored under a given scheduler. This is because the tool runs the program in a truly concurrent environment, thus it only gets to see what the thread did on the particular schedule that was followed.

### 2015-05-12 whoop.tex
Worked with Pantazis on the theory part of the paper and the new proposed structure of the implementation section.

@@

\medskip\noindent\textbf{Data races } \ADComment{We now need to define
what a data race is.  This could be done using happens-before.  I am undecided as to whether
we should refer back to the happens-before material in the background section, or whether
we should move that material here, so that we explain what happens-before means, and then
explain what a data races is.  My preference is for the latter: doing it all here.}

@@

\medskip\noindent\textbf{Shared state abstraction }
\ADComment{Explain why the shared variables are gone.}

\medskip\noindent\textbf{Sequentialisation }
\ADComment{Show what we actually check.}

\ADComment{Now state the theoretical result, paying attention to termination.}

@@

>>>>

\ADComment{Top level question about this section: is the motivation super-clear?  Is it obvious why we go for the pairwise approach?}

@@

@@

\medskip\noindent\textbf{Shared state abstraction }
Recall that while our aim is to perform race analysis for pairs of threads, we must be sure to account
for possible side-effects due to other threads that are running concurrently.  The instrumentation
of Figure~\ref{fig:instrumentation} achieves this via \emph{nondeterminism}: when reading from a shared variable $s$,
a nondeterministic value is returned.  This is reflected by the use of a $\havockeyword$ command,
which sets its argument to an arbitrary value.  Because all shared state accesses are abstracted in
this fashion, it is possible to completely dispense with the shared variables after instrumentation
has been performed.  As a result, when instrumenting a shared variable write, the effect of the write
is not explicitly modelled.

\medskip\noindent\textbf{Sequentialisation }
The psuedocode of Figure~\ref{fig:sequentialization} shows the sequential program that we
analyse in order to prove race-freedom for a pair of thread templates $T$ and $U$.
Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding
$\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows.
First, the read, write and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are
initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and
$\lockset{s}{j}$ are initialized to the full set of mutexes, $\mutexes$.  The main procedures
of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and
$\mainforthread{\instrument{U}{j}}$ are then executed in turn.  Finally, an assertion checks
for consistent use of mutexes: if a shared variables is written during execution of $\instrument{T}{i}$
and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$
and $\lockset{s}{j}$ must contain a common mutex.

\begin{figure}
\begin{tabular}{l}

$\currentlockset{i} = \emptyset;$ $\hasbeenread{i} = \emptyset;$ $\hasbeenwritten{i} = \emptyset;$ \\
$\currentlockset{j} = \emptyset;$ $\hasbeenread{j} = \emptyset;$ $\hasbeenwritten{j} = \emptyset;$ \\
\texttt{for} $s \in \sharedvars$ \texttt{do} $\lockset{s}{i} = \mutexes;$ $\lockset{s}{j} = \mutexes;$ \medskip
\\

$\mainforthread{\instrument{T}{i}}();$ \\
$\mainforthread{\instrument{U}{j}}();$ \medskip\\

\texttt{assert} $\forall s \in \sharedvars \; .$ \\

$\quad s \in \hasbeenwritten{i} \cap (\hasbeenread{j} \cup \hasbeenwritten{j}) \vee s \in \hasbeenwritten{j} \cap (\hasbeenread{i} \cup \hasbeenwritten{i}) \implies$ \\

$\quad\quad \lockset{s}{i} \cap \lockset{s}{j} \neq \emptyset;$ \\

\end{tabular}
\caption{The sequential program to be analysed in order to prove race-freedom for a pair of thread templates}
\label{fig:sequentialization}
\end{figure}

\medskip\noindent\textbf{Soundness }
We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct,
it is impossible for a thread executing template $T$ to race with a thread executing template $U$,
under the assumption that the threads are guaranteed to terminate.
Let us assume that the program of Figure~\ref{fig:sequentialization} is correct,
and suppose (by way of contradition) that a thread executing $T$ can in fact race with a thread
executing $U$, on some shared variable $s$.  By our hypothesis that the program
of Figure~\ref{fig:sequentialization} is correct, and that the threads terminate,
the assertion checked at the end of the program guarantees at least one mutex, $m$ say,
belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset
(and according to the manner in which shared accesses are instrumented in Figure~\ref{ }),
this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$.
As a result, $m$ must be unlocked and locked between the two accesses, which
contradicts that the pair of accesses is racing.

We require the threads to terminate becauase in the presence of non-termination the assertion
at the end of Figure~\ref{fig:sequentialization} may not be reached.  The termination analysis
problem for device drivers has been widely studied, notably in the context of the Terminator tool~\cite{cook2006termination}.
In the remainder of the paper we do not consider termination issues, assuming that the drivers we analyse
in our experimental evaluation (Section~\ref{evaluation}) are terminating.

@@

\ADComment{Acknowledge very clearly that the simple concurrent programming model of Section~\ref{...} is deliberately idealistic.  In practice, we do not know upfront what the entry points for a driver are, we don't have named set of locks, and rather than a given set of named shared variables, we have arbitrary memory accesses via pointers.  We now explain how we have taken the conceptual ideas from Section~\ref{...} and used them to build a practical tool.

\ADComment{Explain briefly the architecture, and specifically introduce Chauffer, SMACK and Boogie.}

\ADComment{Explain how we identify entry points (Chauffer)}

\ADComment{Explain how pointers get handled (SMACK) -- be brief, talk about memory regions, don't show Boogie code.}

\ADComment{Explain how we identify mutexes (inside our Boogie-based analysis)}

\ADComment{Explain how we avoid quantifiers in performing race analysis}

\ADComment{Briefly talk about how we tried inlining and do summarisation instead}

\ADComment{Discuss optimizations and assumptions.  Bring these to before the Corral stuff.}

\ADComment{Note, unrelated to structure: put mention of interrupts in limitations}
